---
title: "PPOL 564 | Week 6 | Breakout"
subtitle: "Manipulating Conflict Event Data" 
output: 
  html_document:
    df_print: paged
    css: async-page-style.css
    highlight: breezedark
    theme: flatly
---

```{r,include = F}
knitr::opts_chunk$set(warning=F,error=F,message = F)
```


# Overview

In this notebook, we'll hone our data manipulation skills by examining conflict event data generated by the [Armed Conflict Location & Event Data Project (ACLED)](https://acleddata.com/#/dashboard). The aim is to practice some of the data manipulation functions covered in the lecture. 

# Data 

ACLED is a "disaggregated data collection, analysis, and crisis mapping project. ACLED collects the dates, actors, locations, fatalities, and modalities of all reported political violence and protest events across Africa, South Asia, Southeast Asia, the Middle East, Central Asia and the Caucasus, Latin America and the Caribbean, and Southeastern and Eastern Europe and the Balkans." For this exercise, we'll focus just on the data pertaining to _Africa_. For more information regarding these data, please consult the [ACLED methodology](https://acleddata.com/resources/methodology/).

<br>

<a href="acled_africa.csv" download><strong>Download the ACLED data</strong></a>.

<br>

```{r,echo=F,warning=F,message=F,commment=F}
readr::read_csv("lectures/week_06/breakout/acled_africa.csv")
```

<br>
<br>

# Know your data

To understand a dataset, one needs to ask a lot of questions of it. To get a better feel for the ACLED data, let's explore the following questions:

<br>
<br>

#### 1. What's the temporal coverage of the data (i.e how many years does it span)?

<br>
<br>

#### 2. How many events are recorded for each country? 

<br>
<br>

#### 3. How many events are recorded for each year?

<br>
<br>


#### 4. What's the most common event type in the data?

<br>
<br>


#### 5. Which countries had the highest number of reported fatalities? 

<br>
<br>

