---
output: 
  html_document:
    css: style.css
    theme: flatly
    highlight: default
    toc_depth: 3
    toc: true
    toc_collapsed: true
    toc_float: true
---

```{r,include=F}
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Data Science I: Foundations</h1></center>
<center><h2>PPOL564-01</h2></center>
<center>**Fall 2020**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}


## Course Schedule

<br>

```{r,echo=F,message=F,results='asis'}
read_csv('Syllabus/Fall-2020-Timeline.csv') %>%
  # select(-`Class No.`,-CD,-Assignment) %>% 
  filter(Topic != "No Class") %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  # kable(., "latex", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                # full_width = T,font_size = 11
                full_width = T,font_size =  18
                ) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '2in') %>% 
  column_spec(3,width = '3in') %>%
  column_spec(4,width = '2in') 
```

<br>
<br>
<br>
<hr>
<br>


## Syllabus

<br>
<br>

- Access an [**Online Version**](syllabus/PPOL564-Syllabus.html)
- [**Download**](syllabus/PPOL564-Syllabus.pdf) a `.pdf` version locally. 

<br>
<br>
<br>
<hr>
<br>

## Readings

<br>
<br>

#### List of all [required reading](readings/reading-list.html).

<br>
<br>
<br>
<hr>
<br>



## Contact

<br>

#### [Eric Dunford](www.ericdunford.com) (Professor)

- **Office**: 404 Old North
- **Office Hours**: Tuesdays 3pm to 5pm
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### Eric LaRose (Teaching Assistant)

- **Office Hours**: by appointment
- **Email**: [eml119@georgetown.edu](mailto:eml119@georgetown.edu)

<br>
<br>
<br>
<hr>
<br>

# Lecture Materials
_Lecture materials will go live approximately one week prior to the scheduled meeting date._

<br><br>

### Week 1: _Introductions, Installations, and IDEs_

<br><br>

### Week 2: _Version Control, Workflow, and Reproducibility_

<br><br>

### Week 3: _Introduction to Object-Oriented Programming in Python_

<br><br>

### Week 4: _Introduction to Algorithms_

<br><br>

### Week 5: _From Nested Lists to Data Frames_

<br><br>

### Week 6: _Data Wrangling in Python_

<br><br>

### Week 7: _Data Visualization and Exploration_

<br><br>

### Week 8: _Scraping and Grabbing: Drawing from (Un-)Structured Data Sources_

<br><br>

### Week 9: _Introduction to Statistical Learning_

<br><br>

### Week 10: _Linear Regression_

<br><br>

### Week 11: _Gradient Descent _

<br><br>

### Week 12: _Introduction to Non-Parametric Approaches to Supervised Learning_

<br><br>

### Week 13: _Peeking inside the Black Box: Introduction to Interpretable Machine Learning_

<br><br><br><br>

<hr>