---
pagetitle: PPOL564 - DS1 - Foundations
output: 
  html_document:
    css: style.css
    theme: flatly
    highlight: breezedark
    toc_depth: 3
    toc: true
    toc_collapsed: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r,include=F}
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```

<center>

<h1>

Data Science I: Foundations

</h1>

</center>

<center>

<h2>

PPOL564-01

</h2>

</center>

<center>

**Fall 2021**

</center>

<center>

**Georgetown University**

</center>

<br>

# Course Outline {.tabset .tabset-fade .tabset-pills}

## Course Schedule

<br>

```{r,echo=F,message=F,warning=F,results='asis'}
read_csv('Syllabus/Fall-2021-Timeline.csv') %>%
  # select(-`Class No.`,-CD,-Assignment) %>% 
  filter(Topic != "No Class") %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l','l','c')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size =  20) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '3in') %>% 
  column_spec(4,width = '2in') %>% 
  column_spec(5,width = '1in')
```

<br> <br> <br>

<hr>

<br>

## Virtual Classroom

<br>

#### [**Virtual Zoom Classroom**](https://georgetown.zoom.us/j/98660441600) (If we need to meet virtually)

<br>

#### [**Virtual Zoom Office Hours**](https://georgetown.zoom.us/j/96444963432) (Mondays/Wednesdays 9am - 10am)

<br>

Recurrent Zoom link can also be found on Canvas. If the link breaks or
does not function properly, please check the `#general` channel on Slack
for information regarding the new link. If there is no message regarding
a new link, please contact the professor and/or TA via Slack. All
synchronous lecture material will be recorded.

<br> <br> <br>

<hr>

<br>

## Syllabus

<br> <br>

#### Access an [**online version**](syllabus/PPOL564-Syllabus.html)/[**`.pdf` version**](syllabus/PPOL564-Syllabus.pdf).

<br> <br> <br>

<hr>

<br>

## Readings

<br> <br>

#### List of all [**required reading**](readings/reading-list.html).

<br> <br> <br>

<hr>

<br>

## Installations {.tabset}

<br>

Throughout the semester, the instructor will use the commandline and
many different IDEs when coding in Python or using Git. The following
lists those different software and provides guidance on installation.
*If you run into issues, please reach out to the Teaching Assistant for
assistance.*

<br>

### Commandline

<br>

At times, we'll use a unix-based commandline. The commandline will
feature into our discussion on using `git` and also running Python
programs. If you use a Mac or a Linux operating system, then a
functioning commandline comes with your operating system. For Apple
machines, this is the **Terminal**.

For Windows (specifically Windows 10), you can enable Linux Bash shell.
The following offers a
[tutorial](https://www.howtogeek.com/249966/how-to-install-and-use-the-linux-bash-shell-on-windows-10/)
on how to do this. If you're using a version of Windows that pre-dates
version 10, then [Git
Bash](https://www.atlassian.com/git/tutorials/git-bash#:~:text=Git%20Bash%20is%20an%20application,operating%20system%20through%20written%20commands.)
offers a program will allow you to use `git` commands from your windows
machine.

Finally, you'll notice that my terminal will have a slightly different
look than the one on your machine. This is because I'm using ["Oh My
Zsh"](https://ohmyz.sh/) which is open-source software that allows me to
customize my commandline. The above link offers everything you'd need to
installing Oh My Zsh on your machine.

<br>

<hr>

<br>

### Python3

<br>

We'll use Python3 throughout this course. Below are instructions for
downloading Python3 using commandline packages manager (`Homebrew` for
mac, `Chocolatey` for windows).

<br>

-   [Installing Python3 using Homebrew on a Mac
    (Apple)](https://docs.python-guide.org/starting/install3/osx/)

-   [Installing Python3 using Chocolatey on a PC
    (Windows)](https://python-docs.readthedocs.io/en/latest/starting/install3/win.html)

<br>

An alternative way to install Python3 is to download an [**Anaconda**
distribution](https://www.anaconda.com/products/individual). The
instructor will use `pip` rather than `conda` in the instruction for
downloading Python modules. These are simply two ways of downloading and
managing open-source software packages. <u>Choose which ever works best
for you</u>.

<br>

<hr>

<br>

### Jupyter Notebook

<br>

Once you have Python3 on your computer, you can install a Jupyter
Notebook. If you downloaded Python3 using Anaconda, then Jupyter
Notebook comes with the distribution and requires no further
installation on your part. If you install Python3 using
`Homebrew`/`Chocolately`, you can install Jupyter notebook running the
following code using your commandline.

```{bash,eval=F}
pip install notebook
```

You can then activate a Jupyter Notebook from the commandline by typing:

```{bash,eval=F}
jupyter notebook
```

If you've installed Python using Anaconda, the distribution provides a
click-able icon to fire up a Jupyter Notebook. The advantage of using
the commandline, however, is that you can set the working directory
prior to firing up a notebook. This will allow you to work within a
specific project folder more easily.

<br>

<hr>

<br>

### Atom + `hydrogen`

<br>

[**Atom**](https://atom.io/) is a hack-able text editor built by Github.
The following are [instruction on how to install
Atom](https://flight-manual.atom.io/getting-started/sections/installing-atom/)
on your machine.

Atom allows you to install open-source packages that provide additional
functionality. The following packages will help you as you use Atom to
program in Python. Of these,
[`Hydrogen`](https://atom.io/packages/hydrogen) is the most important.
It'll allow you to use a Jupyter kernel from within Atom to evaluate
code.

```{bash,eval=F}
Hydrogen@2.16.3
Zen@0.18.0
advanced-open-file@0.16.8
atom-beautify@0.33.4
atom-clock@0.1.18
atom-html-preview@0.2.6
atom-language-r@1.4.8
atom-material-syntax@1.0.8
atom-material-syntax-light@0.4.6
atom-material-ui@2.1.3
atom-path-intellisense@1.2.2
atom-python-virtualenv@1.0.4
atom-todoist@2.0.0
auto-update-packages@1.0.1
autocomplete-R@0.6.0
autocomplete-latex-cite@0.3.5
autocomplete-modules@2.3.0
autocomplete-python@1.17.0
autocomplete-sql@0.5.0
browser-plus@0.0.98
color-picker@2.3.0
data-explorer@0.7.0
docblock-python@0.19.1
file-icons@2.1.47
fix-indent-on-paste@0.1.1
fold-comments@0.6.0
git-log@0.4.1
hey-pane@1.2.0
hydrogen-cell-separator@0.4.1
indent-guide-improved@1.4.13
jupyter-notebook@0.0.10
kite@0.206.0
language-latex@1.2.0
language-weave@0.7.2
latex@0.50.2
latex-tree@0.5.0
latexer@0.3.0
minimap@4.39.14
oceanic-next@1.0.0
pdf-view@0.73.0
platformio-ide-terminal@2.10.1
project-manager@3.3.8
python-indent@1.2.6
quick-query-sqlite@0.4.1
reindent@1.5.0
scroll-through-time@0.3.1
simple-drag-drop-text@0.5.0
symbols-tree-view@0.14.0
todo-show@2.3.2
typewriter@0.8.0
wordcount@3.2.0
```

To install any one of these packages from the commandline, type:

```{bash,eval = F}
# apm == "Atom package manager"
apm install <package-name>
# For example
apm install Hydrogen@2.16.3
```

There is also a dedicated package manager built into Atom which you can
use to download and install new packages. Open Atom then
`Settings > Install` and type the package name.

#### Troubleshooting Hydrogen/Atom Setup

Several students have had issues arise in getting Hydrogen to properly
run on their machines. Particularly, after following the installation
instructions for Atom and Hydrogen, many people find that upon trying to
run Python code, they either (1) receive an error message stating that
`"no kernel for language Python found"` (or something similar), or (2)
they are able to connect to a Python kernel but upon trying to run code,
nothing happens (they may or may not receive error messages associated
with that.

If you encounter this issue, we suggest trying the following solutions
in order until one of the solutions works. If you have tried all three
possible solutions and are still not able to properly run Python code in
Hydrogen/Atom, please contact the teaching assistants (either by Slack,
email, or setting up a Calendly appointment).

<br>

**Solution 1**

Open the command line and run the following two commands:

    python3 -m pip install ipykernel
    python3 -m ipykernel install --user

Then restart Atom and try running Python code.

<br>

**Solution 2**

Uninstall Hydrogen on Atom by opening Atom, click `"Install a Package"`,
and search for Hydrogen in the search bar. Click `"Uninstall"`. Once
Hydrogen has finished uninstalling, search Hydrogen again and hit
`"Install"`. Once Hydrogen has finished re-installing, restart Atom and
try running Python code.

<br>

**Solution 3**

Add the following paths to your list of environmental variables using
the command line. Note that exact file paths will need to be adjusted
slightly depending on your machine and operating system.

    C:\Anaconda3
    C:\Anaconda3\Scripts
    C:\Anaconda3\Library\bin

Once these have been added to the list of environmental variables,
restart Atom and try running Python code.

<br>

<hr>

<br>

### RStudio + `reticulate`

<br>

In your classes that are focused on using `R`, `RStudio` will be your
main IDE. However, `RStudio` isn't just for `R`. It can handle a number
of different languages. We can use Python in `RStudio` using the
`reticulate` package. We'll talk about some of the advantages for doing
this in class, but for now, let's cover installation.

To install `RStudio`, [download from the following
link](https://www.rstudio.com/products/rstudio/download/) (make sure to
scroll all the way to the bottom).

`reticulate` is an R package that allows one run a Python REPL in the R
console. In addition, it allows one to read in and use Python code, and
pass data between R and and Python. The following provides [instructions
on installing `reticulate`](https://rstudio.github.io/reticulate/).

**Note**: If you have multiple versions of Python on your computer,
reticulate can get confused with regard to which version it is
referencing. The [following
article](https://rstudio.github.io/reticulate/articles/versions.html)
covers these issues. The best way to resolve this issue is by creating a
`.Rprofile` file that sends instructions regarding the specific version
of Python you wish to use.

> "By setting the value of the RETICULATE_PYTHON environment variable to
> a Python binary. Note that if you set this environment variable, then
> the specified version of Python will always be used (i.e. this is
> prescriptive rather than advisory). To set the value of
> RETICULATE_PYTHON, insert Sys.setenv(RETICULATE_PYTHON = PATH) into
> your project's .Rprofile, where PATH is your preferred Python binary."

<br>

<hr>

<br>

### Other Software

<br>

Here is an overview of other text editors that are popular for
programming in Python, which you won't see featured in this course. Note
I'm agnostic on whatever you use to learn Python and some find that
different set ups work better for them. If one of these setups works
better for you, I encourage you to use it (and tell me about how it
went)!

<br>

-   [Sublime Text](https://www.sublimetext.com/)
-   [PyCharm](https://www.jetbrains.com/pycharm/)
-   [Spyder](https://www.spyder-ide.org/)
-   [Rodeo](https://rodeo.yhat.com/)
-   [Visual Studio](https://code.visualstudio.com/docs/languages/python)

<br>

<hr>

<br>

## Project

<br>

Data science is an applied field and therefore, it is important that you
understand how to conduct a complete analysis from collecting data, to
cleaning and analyzing it, to presenting your findings. Toward this end,
students are required to complete an independent data science project,
*applying concepts learned throughout the course*. The project is
composed of three parts: a 2 page project proposal, an in-class
presentation, and a 12-page project report.

***More information regarding the final project will be circulated
during class on Week 8***

<br>

<center><strong>[Final Project Overview](project/project_overview.pdf)</strong></center>

<!-- <br> -->

<!-- <center><strong>[Presentation Rubric](project/project_presentation_assessment_sheet.pdf)</strong></center> -->

<!-- <br> -->

<!-- <center><strong>[Final Report Rubric](project/final_project_rubric.pdf)</strong></center> -->

<!-- <br> -->

<br> <br> <br>

<hr>

<br>

## Contact

<br>

#### [Eric Dunford](http://ericdunford.com/) (Professor)

-   **Office**: 404 Old North
-   **Office Hours**: Mondays & Wednesdays 9am to 10am (Office hours will be held virtually via Zoom. See `Virtual Classroom` tab for link)
-   **Email**:
    [eric.dunford\@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### [Maddie Pickens](https://maddieclairepickens.com/) (Teaching Assistant)

-   **Office Hours**: [Book an appointment via
    Calendly](https://calendly.com/maddiepickens)
-   **Email**: [mp1595\@georgetown.edu](mailto:mp1595@georgetown.edu)

<br>

#### [Chandler Dawson](https://www.linkedin.com/in/chandler-dawson-586615114/) (Teaching Assistant)

-   **Office Hours**: [Book an appointment via
    Calendly](https://calendly.com/cad162)
-   **Email**: [cad162\@georgetown.edu](mailto:cad162@georgetown.edu)

<br>

#### Slack

The best way to reach the TA/Professor is via the class Slack channel
(PPOL-564-Fall-2021). Please click on the [Class Slack Channel
Invite](https://join.slack.com/t/ppol-564-fall-2021/shared_invite/zt-ukj26377-a00MPhjVTkh2rs8VSTrmzw)
to join the class work-space.

<br> <br> <br>

<hr>

<br>

# Lecture Materials

*Asynchronous lecture materials will go live approximately one week
prior to the scheduled synchronous meeting date.*

<br><br>

### **Week 1: Choosing your Poison**<br>[*Introductions, Installations, and IDEs*](lectures/week_01/asynchronous_lecture_material/week01-async-materials.html)

<!-- _Introductions, Installations, and IDEs_ -->

<br><br>

### **Week 2: Time Travel and Other Necessities**<br>[*Version Control, Workflow, and Reproducibility*](lectures/week_02/asynchronous_lecture_material/week-02-async-materials.html)

<!-- _Version Control, Workflow, and Reproducibility_ -->

<br><br>

### **Week 3: Learning Parseltongue**<br>[*Object-Oriented Programming in Python*](lectures/week_03/asynchronous_lecture_material/week-03-async-material.html)

<!-- _Object-Oriented Programming in Python_ -->

<br><br>

### **Week 4: On Time and Space**<br>[_Introduction to Algorithms_](lectures/week_04/week-04-async-material.html)

<!-- *Introduction to Algorithms* -->

<br><br>

### **Week 5: Long Live the Data Frame**<br>[_From Nested Lists to Data Frames_](lectures/week_05/week-05-async-material.html)

<!-- *From Nested Lists to Data Frames* -->

<br><br>

### **Week 6: Modern Snake Charming**<br>[_Approaches to Data Manipulation in Python_](lectures/week_06/week-06-async-material.html)

<!-- *Approaches to Data Manipulation in Python* -->

<br><br>

### **Week 7: Interrogation Techniques**<br>[_Data Visualization and Exploration_](lectures/week_07/week-07-async-material.html)

<!-- *Data Visualization and Exploration* -->

<br><br>

### **Week 8: Automated Heists**<br>[_Drawing from (Un-)Structured Data Sources_](lectures/week_08/week-08-async-material.html)

<!-- *Drawing from (Un-)Structured Data Sources* -->

<br><br>

### **Week 9: The Signal and the Noise**<br>[_Introduction to Statistical Learning_](lectures/week_09/week-09-async-material.html)

<!-- *Introduction to Statistical Learning* -->

<br><br>

### **Week 10: Casting Shadows in** $N$-Dimensions<br>[_Continuous Outcomes and Linear Regression_](lectures/week_10/week-10-async-material.html)

<!-- *Continuous Outcomes and Linear Regression* -->

<br><br>

### **Week 11: Hot Dog, Not Hot Dog**<br>[_Probability, Bayes Theorem, and Classification_](lectures/week_11/week-11-async-material.html)

<!-- *Probability, Bayes Theorem, and Classification* -->

<br><br>

### **Week 12: Trees and Neighbors**<br>[_Algorithmic Approaches to Supervised Learning_](lectures/week_12/week-12-async-material.html)

<!-- *Algorithmic Approaches to Supervised Learning* -->

<br><br>

### **Week 13: Peeking inside the Black Box**<br>[_Interpretable Machine Learning_](lectures/week_13/week-13-async-material.html)

<!-- *Interpretable Machine Learning* -->

<br><br><br><br>

<hr>
