---
pagetitle: PPOL564 - DS1 - Foundations
output: 
  html_document:
    css: style.css
    theme: flatly
    highlight: breezedark
    toc_depth: 3
    toc: true
    toc_collapsed: true
    toc_float: true
---

```{r,include=F}
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Data Science I: Foundations</h1></center>
<center><h2>PPOL564-01</h2></center>
<center>**Fall 2020**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}


## Course Schedule

<br>

```{r,echo=F,message=F,warning=F,results='asis'}
read_csv('Syllabus/Fall-2020-Timeline.csv') %>%
  # select(-`Class No.`,-CD,-Assignment) %>% 
  filter(Topic != "No Class") %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l','l','c')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size =  20) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '3in') %>% 
  column_spec(4,width = '2in') %>% 
  column_spec(5,width = '1in')
```

<br>
<br>
<br>
<hr>
<br>

## Virtual Classroom


<br> 


#### [**Virtual Zoom Classroom**](https://georgetown.zoom.us/j/98660441600)
  
<br>
  
#### [**Virtual Zoom Office Hours**](https://georgetown.zoom.us/j/96444963432) (Tuesdays 9am - 11am)

  
<br>

Recurrent Zoom link can also be found on Canvas. If the link breaks or does not function properly, please check the `#general` channel on Slack for information regarding the new link. If there is no message regarding a new link, please contact the professor and/or TA via Slack. All synchronous lecture material will be recorded.

<br>
<br>
<br>
<hr>
<br>


## Syllabus

<br>
<br>

#### Access an [**online version**](syllabus/PPOL564-Syllabus.html)/[**`.pdf` version**](syllabus/PPOL564-Syllabus.pdf).

<br>
<br>
<br>
<hr>
<br>

## Readings

<br>
<br>

#### List of all [**required reading**](readings/reading-list.html).

<br>
<br>
<br>
<hr>
<br>

## Installations {.tabset}

<br>

Throughout the semester, the instructor will use the commandline and many different IDEs when coding in Python or using Git. The following lists those different software and provides guidance on installation. _If you run into issues, please reach out to the Teaching Assistant for assistance._ 

<br>

### Commandline

<br>

At times, we'll use a unix-based commandline. The commandline will feature into our discussion on using `git` and also running Python programs. If you use a Mac or a Linux operating system, then a functioning commandline comes with your operating system. For Apple machines, this is the **Terminal**. 

For Windows (specifically Windows 10), you can enable Linux Bash shell. The following offers a [tutorial](https://www.howtogeek.com/249966/how-to-install-and-use-the-linux-bash-shell-on-windows-10/) on how to do this. If you're using a version of Windows that pre-dates version 10, then [Git Bash](https://www.atlassian.com/git/tutorials/git-bash#:~:text=Git%20Bash%20is%20an%20application,operating%20system%20through%20written%20commands.) offers a program will allow you to use `git` commands from your windows machine. 

Finally, you'll notice that my terminal will have a slightly different look than the one on your machine. This is because I'm using ["Oh My Zsh"](https://ohmyz.sh/) which is open-source software that allows me to customize my commandline. The above link offers everything you'd need to installing Oh My Zsh on your machine. 

<br>
<hr>
<br>

### Python3 

<br>

We'll use Python3 throughout this course. Below are instructions for downloading Python3 using commandline packages manager (`Homebrew` for mac, `Chocolatey` for windows). 

<br>

- [Installing Python3 using Homebrew on a Mac (Apple)](https://docs.python-guide.org/starting/install3/osx/)

- [Installing Python3 using Chocolatey on a PC (Windows)](https://python-docs.readthedocs.io/en/latest/starting/install3/win.html)

<br>

An alternative way to install Python3 is to download an [**Anaconda** distribution](https://www.anaconda.com/products/individual). The instructor will use `pip` rather than `conda` in the instruction for downloading Python modules. These are simply two ways of downloading and managing open-source software packages. <u>Choose which ever works best for you</u>. 


<br>
<hr>
<br>

### Jupyter Notebook

<br> 

Once you have Python3 on your computer, you can install a Jupyter Notebook. If you downloaded Python3 using Anaconda, then Jupyter Notebook comes with the distribution and requires no further installation on your part. If you install Python3 using `Homebrew`/`Chocolately`, you can install Jupyter notebook running the following code using your commandline.

```{bash,eval=F}
pip install notebook
```
You can then activate a Jupyter Notebook from the commandline by typing:

```{bash,eval=F}
jupyter notebook
```

If you've installed Python using Anaconda, the distribution provides a click-able icon to fire up a Jupyter Notebook. The advantage of using the commandline, however, is that you can set the working directory prior to firing up a notebook. This will allow you to work within a specific project folder more easily. 

<br>
<hr>
<br>

### Atom + `hydrogen`

<br>

[**Atom**](https://atom.io/) is a hack-able text editor built by Github. The following are [instruction on how to install Atom](https://flight-manual.atom.io/getting-started/sections/installing-atom/) on your machine. 

Atom allows you to install open-source packages that provide additional functionality. The following packages will help you as you use Atom to program in Python. Of these, [`Hydrogen`](https://atom.io/packages/hydrogen) is the most important. It'll allow you to use a Jupyter kernel from within Atom to evaluate code.

```{bash,eval=F}
├── Hydrogen@2.14.4
├── atom-beautify@0.33.4
├── atom-language-r@1.4.8
├── atom-material-syntax@1.0.8
├── atom-material-syntax-light@0.4.6
├── atom-material-ui@2.1.3
├── auto-update-packages@1.0.1
├── autocomplete-R@0.6.0
├── autocomplete-modules@2.3.0
├── autocomplete-python@1.16.0
├── color-picker@2.3.0
├── docblock-python@0.19.0
├── file-icons@2.1.42
├── fix-indent-on-paste@0.1.1
├── fold-comments@0.6.0
├── git-log@0.4.1
├── hey-pane@1.2.0
├── hydrogen-cell-separator@0.4.1
├── language-weave@0.7.2
├── minimap@4.29.9
├── pdf-view@0.72.0
├── platformio-ide-terminal@2.10.0
├── python-indent@1.2.6
├── reindent@1.5.0
├── simple-drag-drop-text@0.5.0
├── symbols-tree-view@0.14.0
└── wordcount@3.2.0
```

To install any one of these packages from the commandline, type:
```{bash,eval = F}
# apm == "Atom package manager"
apm install <package-name>
```

There is also a dedicated package manager built into Atom which you can use to download and install new packages. Open Atom then `Settings > Install` and type the package name. 

#### Troubleshooting Hydrogen/Atom Setup

Several students have had issues arise in getting Hydrogen to properly run on their machines. Particularly, after following the installation instructions for Atom and Hydrogen, many people find that upon trying to run Python code, they either (1) receive an error message stating that `"no kernel for language Python found"` (or something similar), or (2) they are able to connect to a Python kernel but upon trying to run code, nothing happens (they may or may not receive error messages associated with that.

If you encounter this issue, we suggest trying the following solutions in order until one of the solutions works. If you have tried all three possible solutions and are still not able to properly run Python code in Hydrogen/Atom, please contact Eric LaRose (either by Slack, email, or setting up a Calendly appointment). 

<br>

**Solution 1**

Open the command line and run the following two commands:

```
python3 -m pip install ipykernel
python3 -m ipykernel install --user
```

Then restart Atom and try running Python code.

<br>

**Solution 2**

Uninstall Hydrogen on Atom by opening Atom, click `"Install a Package"`, and search for Hydrogen in the search bar. Click `"Uninstall"`. Once Hydrogen has finished uninstalling, search Hydrogen again and hit `"Install"`. Once Hydrogen has finished re-installing, restart Atom and try running Python code. 

<br>

**Solution 3**

Add the following paths to your list of environmental variables using the command line. Note that exact file paths will need to be adjusted slightly depending on your machine and operating system.

```
C:\Anaconda3
C:\Anaconda3\Scripts
C:\Anaconda3\Library\bin
```


Once these have been added to the list of environmental variables, restart Atom and try running Python code. 


<br>
<hr>
<br>

### RStudio + `reticulate`

<br>

In your classes that are focused on using `R`, `RStudio` will be your main IDE. However, `RStudio` isn't just for `R`. It can handle a number of different languages. We'll Python in `RStudio` using the `reticulate` package. We'll talk about some of the advantages for doing this in class, but for now, let's cover installation.

To install `RStudio`, [download from the following link](https://www.rstudio.com/products/rstudio/download/) (make sure to scroll all the way to the bottom).

`reticulate` is an R package that allows one run a Python REPL in the R console. In addition, it allows one to read in and use Python code, and pass data between R and and Python. The following provides [instructions on installing `reticulate`](https://rstudio.github.io/reticulate/).

**Note**: If you have multiple versions of Python on your computer, reticulate can get confused with regard to which version it is referencing. The [following article](https://rstudio.github.io/reticulate/articles/versions.html) covers these issues. The best way to resolve this issue is by creating a `.Rprofile` file that sends instructions regarding the specific version of Python you wish to use. 

> "By setting the value of the RETICULATE_PYTHON environment variable to a Python binary. Note that if you set this environment variable, then the specified version of Python will always be used (i.e. this is prescriptive rather than advisory). To set the value of RETICULATE_PYTHON, insert Sys.setenv(RETICULATE_PYTHON = PATH) into your project’s .Rprofile, where PATH is your preferred Python binary."
 


<br>
<hr>
<br>

### Other Software

<br>

Here is an overview of other text editors that are popular for programming in Python, which you won't see featured in this course. Note I'm agnostic on whatever you use to learn Python and some find that different set ups work better for them. If one of these setups works better for you, I encourage you to use it (and tell me about how it went)!

<br>

- [Sublime Text](https://www.sublimetext.com/)
- [PyCharm](https://www.jetbrains.com/pycharm/)
- [Spyder](https://www.spyder-ide.org/)
- [Rodeo](https://rodeo.yhat.com/)
- [Visual Studio](https://code.visualstudio.com/docs/languages/python)

<br>
<hr>
<br>


## Project

<br>

Data science is an applied field and therefore, it is important that you understand how to conduct a complete analysis from collecting data, to cleaning and analyzing it, to presenting your findings. Toward this end, students are required to complete an independent data science project, _applying concepts learned throughout the course_. The project is composed of three parts: a 2 page project proposal, an in-class presentation, and a 12-page project report.

**_More information regarding the final project will be circulated during class on Week 8_**

<br>
<br>
<br>
<hr>
<br>


## Contact

<br>

#### [Eric Dunford](http://ericdunford.com/) (Professor)

- **Office**: Bedroom (formerly 404 Old North)
- **Office Hours**: Tuesdays 9am to 11am (See `Virtual Classroom` tab for link)
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### Eric LaRose (Teaching Assistant)

- **Office Hours**: [Book an appointment via Calendly](https://calendly.com/eml119/)
- **Email**: [eml119@georgetown.edu](mailto:eml119@georgetown.edu)

<br>

#### Slack

The best way to reach the TA/Professor is via the class Slack channel (PPOL-564-Fall-2020). Please click on the [Class Slack Channel Invite](https://join.slack.com/t/georgetown-jt44872/shared_invite/zt-gntolqrn-BiCH6ChQTKKODMyb5PdRAw) to join the class work-space. 

<br>
<br>
<br>
<hr>
<br>

# Lecture Materials
_Asynchronous lecture materials will go live approximately one week prior to the scheduled synchronous meeting date._

<br><br>

### **Week 1: Choosing your Poison**<br>[_Introductions, Installations, and IDEs_](lectures/week_01/asynchronous_lecture_material/week01-async-materials.html)

<br><br>

### **Week 2: Time Travel and Other Necessities**<br>[_Version Control, Workflow, and Reproducibility_](lectures/week_02/asynchronous_lecture_material/week-02-async-materials.html)

<br><br>

### **Week 3: Learning Parseltongue**<br>[_Object-Oriented Programming in Python_](lectures/week_03/asynchronous_lecture_material/week-03-async-material.html)

<br><br>

### **Week 4: On Time and Space**<br>[_Introduction to Algorithms_](lectures/week_04/week-04-async-material.html)

<br><br>

### **Week 5: Long Live the Data Frame**<br>[_From Nested Lists to Data Frames_](lectures/week_05/week-05-async-material.html)

<br><br>

### **Week 6: Modern Snake Charming**<br>[_Approaches to Data Manipulation in Python_](lectures/week_06/week-06-async-material.html)

<br><br>

### **Week 7: Interrogation Techniques**<br>[_Data Visualization and Exploration_](lectures/week_07/week-07-async-material.html)

<br><br>

### **Week 8: Automated Heists**<br>[_Drawing from (Un-)Structured Data Sources_](lectures/week_08/week-08-async-material.html)

<br><br>

### **Week 9: The Signal and the Noise**<br>_Introduction to Statistical Learning_

<br><br>

### **Week 10: Casting Shadows in $N$-Dimensions**<br>_Continuous Outcomes and Linear Regression_

<br><br>

### **Week 11: Hot Dog, Not Hot Dog**<br>_Probability, Bayes Theorem, and Classification_

<br><br>

### **Week 12: Trees and Neighbors**<br>_Non-Parametric Approaches to Supervised Learning_

<br><br>

### **Week 13: Peeking inside the Black Box**<br>_Interpretable Machine Learning_

<br><br><br><br>

<hr>