---
pagetitle: PPOL564 - DS1 - Foundations
output: 
  html_document:
    css: style.css
    theme: flatly
    highlight: default
    toc_depth: 3
    toc: true
    toc_collapsed: true
    toc_float: true
---

```{r,include=F}
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Data Science I: Foundations</h1></center>
<center><h2>PPOL564-01</h2></center>
<center>**Fall 2020**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}


## Course Schedule

<br>

```{r,echo=F,message=F,warning=F,results='asis'}
read_csv('Syllabus/Fall-2020-Timeline.csv') %>%
  # select(-`Class No.`,-CD,-Assignment) %>% 
  filter(Topic != "No Class") %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l','l','c')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size =  20) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '3in') %>% 
  column_spec(4,width = '2in') %>% 
  column_spec(5,width = '1in')
```

<br>
<br>
<br>
<hr>
<br>

## Virtual Classroom


<br> 


#### [**Virtual Zoom Classroom**](https://georgetown.zoom.us/j/98660441600)
  
<br>
  
#### [**Virtual Zoom Office Hours**](https://georgetown.zoom.us/j/96444963432) (Tuesdays 9am - 11am)

  
<br>

Recurrent Zoom link can also be found on Canvas. If the link breaks or does not function properly, please check the `#general` channel on Slack for information regarding the new link. If there is no message regarding a new link, please contact the professor and/or TA via Slack. All synchronous lecture material will be recorded.

<br>
<br>
<br>
<hr>
<br>


## Syllabus

<br>
<br>

#### Access an [**online version**](syllabus/PPOL564-Syllabus.html)/[**`.pdf` version**](syllabus/PPOL564-Syllabus.pdf).

<br>
<br>
<br>
<hr>
<br>

## Readings

<br>
<br>

#### List of all [**required reading**](readings/reading-list.html).

<br>
<br>
<br>
<hr>
<br>


## Project

<br>

Data science is an applied field and therefore, it is important that you understand how to conduct a complete analysis from collecting data, to cleaning and analyzing it, to presenting your findings. Toward this end, students are required to complete an independent data science project, _applying concepts learned throughout the course_. The project is composed of three parts: a 2 page project proposal, an in-class presentation, and a 12-page project report.

**_More information regarding the final project will be circulated during class on Week 8_**

<br>
<br>
<br>
<hr>
<br>


## Contact

<br>

#### [Eric Dunford](http://ericdunford.com/) (Professor)

- **Office**: Bedroom (formerly 404 Old North)
- **Office Hours**: Tuesdays 9am to 11am (See `Virtual Classroom` tab for link)
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### Eric LaRose (Teaching Assistant)

- **Office Hours**: [Book an appointment via Calendly](https://calendly.com/eml119/)
- **Email**: [eml119@georgetown.edu](mailto:eml119@georgetown.edu)

<br>

#### Slack

The best way to reach the TA/Professor is via the class Slack channel (PPOL-564-Fall-2020). Please click on the [Class Slack Channel Invite](https://join.slack.com/t/georgetown-jt44872/shared_invite/zt-gntolqrn-BiCH6ChQTKKODMyb5PdRAw) to join the class work-space. 

<br>
<br>
<br>
<hr>
<br>

# Lecture Materials
_Asynchronous lecture materials will go live approximately one week prior to the scheduled synchronous meeting date._

<br><br>

### **Week 1: Choosing your Poison**<br>_Introductions, Installations, and IDEs_

<br><br>

### **Week 2: Time Travel and Other Necessities**<br>_Version Control, Workflow, and Reproducibility_

<br><br>

### **Week 3: Learning Parseltongue**<br>_Object-Oriented Programming in Python_

<br><br>

### **Week 4: On Time and Space**<br>_Introduction to Algorithms and Data Structures_

<br><br>

### **Week 5: Long Live the Data Frame**<br>_From Nested Lists to Data Frames_

<br><br>

### **Week 6: Modern Snake Charming**<br>_Approaches to Data Manipulation in Python_

<br><br>

### **Week 7: Interrogation Techniques**<br>_Data Visualization and Exploration_

<br><br>

### **Week 8: Automated Heists**<br>_Drawing from (Un-)Structured Data Sources_

<br><br>

### **Week 9: The Signal and the Noise**<br>_Introduction to Statistical Learning_

<br><br>

### **Week 10: Casting Shadows in $N$-Dimensions**<br>_Continuous Outcomes and Linear Regression_

<br><br>

### **Week 11: Hot Dog, Not Hot Dog**<br>_Probability, Bayes Theorem, and Classification_

<br><br>

### **Week 12: Trees and Neighbors**<br>_Non-Parametric Approaches to Supervised Learning_

<br><br>

### **Week 13: Peeking inside the Black Box**<br>_Interpretable Machine Learning_

<br><br><br><br>

<hr>